---
title: "Prediction Assignment Writeup"
author: "Lorenzo A Sanchez"
date: "September 15, 2017"
output: html_document
---

## Prediction of Unilateral Dumbbell Biceps Curl exercise forms using Human Activity Recognition (HAR) devices.

#### 1. Background and Objective
The goal of this exercise is to develop a Machine Learning (ML) model using data collected from 6 participants wearing accelerometers on the belt, forearm, arm, and dumbbell that will predict which out of 5 exercise forms was used when performing a unilateral dumbbell bicep curl. Only class A is considered correct form while the other classes are not.

#### 2. Data preparation

##### 2.1 Loading the data and libraries

The read.table statement will also turn "NA", "#DIV/0!", and empty strings to NA's.
```{r warning=FALSE, message=FALSE, error=FALSE}
library(caret); library(rattle); library(gridExtra); library(randomForest)

setwd("C:/Users/lsanchez/Desktop/Coursera Data Science/Course 8 Practical Machine Learning")

training <- read.table("./pml-training.csv", header = TRUE, 
                       sep = ",", na.strings = c("NA","#DIV/0!",""), stringsAsFactors=FALSE)
testing <- read.table("./pml-testing.csv", header = TRUE,
                      sep = ",", na.strings = c("NA","#DIV/0!",""), stringsAsFactors=FALSE)
```

##### 2.2 Clean data

###### 2.2.1 Remove dimensions with Near Zero Variance (NZV) or irrelevant labels
Obtain NZV statistics to then clean data:
```{r warning=FALSE, message=FALSE, error=FALSE}
training_NZV <- nearZeroVar(training, saveMetrics=TRUE)

training_no_NZV <- training[,!(training_NZV$zeroVar | training_NZV$nzv)]

testing_no_NZV <- testing[,!(training_NZV$zeroVar | training_NZV$nzv)]
```

Removed the 36 variables with NZV leaving only 124 variables.

Also remove ID column and raw day and second counts (columns 1, 3, and 4:
```{r warning=FALSE, message=FALSE, error=FALSE}
training_noNZV_noID <- training_no_NZV[,-c(1,3,4)]
testing_noNZV_noID <- testing_no_NZV[,-c(1,3,4)]
```

###### 2.2.2 Remove dimensions where 90% or more of values are NAs
```{r warning=FALSE, message=FALSE, error=FALSE}
NA_percent = vector()

for (i in 1:length(names(testing_noNZV_noID))) {
  NA_percent[[i]] <-  sum(is.na(training_noNZV_noID[,i]))/length(training_noNZV_noID[,i])
}

NA_thereshold <- NA_percent < 0.9
training_noNZV_noID_noNA <- training_noNZV_noID[,NA_thereshold] 

testing_noNZV_noID_noNA <- testing_noNZV_noID[,NA_thereshold] 
```

###### 2.2.3 Convert Date values to Date class
```{r warning=FALSE, message=FALSE, error=FALSE}
training_noNZV_noID_noNA$cvtd_timestamp <-  as.Date(training_noNZV_noID_noNA$cvtd_timestamp, format = "%d/%m/%y")


testing_noNZV_noID_noNA$cvtd_timestamp <-  as.Date(testing_noNZV_noID_noNA$cvtd_timestamp, format = "%d/%m/%y")

training_clean <- training_noNZV_noID_noNA
testing_clean <- testing_noNZV_noID_noNA
```

###### 2.2.4 Center, scale, and normalize numeric values

Centering data will substract mean from value, scaling data will divide values by standard deviation, and normalizing values will transform range of values from 0 to 1.
```{r warning=FALSE, message=FALSE, error=FALSE}
numeric_dims = vector()

for (i in 1:length(names(training_clean))) {
  numeric_dims[i] <- is.numeric(training_clean[,i]) 
}
#Scale
scale_params_training <- preProcess(training_clean[,numeric_dims], method=c("scale"))
training_scaled <- predict(scale_params_training, training_clean[,numeric_dims])

scale_params_testing <- preProcess(testing_clean[,numeric_dims], method=c("scale"))
testing_scaled <- predict(scale_params_testing, testing_clean[,numeric_dims])  

#Center
center_params_training <- preProcess(training_scaled, method=c("center"))
training_centered <- predict(center_params_training, training_scaled)

center_params_testing <- preProcess(testing_scaled, method=c("center"))
testing_centered <- predict(center_params_testing, testing_scaled) 

#Normalize
normalize_params_training <- preProcess(training_centered, method=c("range"))
training_normalized <- predict(normalize_params_training, training_centered)

normalize_params_testing <- preProcess(testing_centered, method=c("range"))
testing_normalized <- predict(normalize_params_testing, testing_centered) 

training_clean[,numeric_dims] <- training_normalized
testing_clean[,numeric_dims] <- testing_normalized
```

##### 2.3 Partition training data into training and validation

The validation partition will be 25% of the original training data.
```{r warning=FALSE, message=FALSE, error=FALSE}

set.seed(888)
inTrain <- createDataPartition(y=training_clean$classe, p=0.75, list=FALSE)
myTraining <- training_clean[inTrain, ]
myTesting <- training_clean[-inTrain, ]
```

#### 3. Train Random Forest model
```{r warning=FALSE, message=FALSE, error=FALSE}
set.seed(888)
RF_model <- train(classe~., method="rf", data=myTraining, trControl=trainControl(method="cv"),
                  number=5, allowParallel=TRUE, importance=TRUE)
print(RF_model)
```
The accurarcy of the trained random forest model on the training data is of 99.73%.

```{r warning=FALSE, message=FALSE, error=FALSE}
training_predict <- predict(RF_model,myTraining)
confusionMatrix(training_predict,myTraining$classe)
```

```{r warning=FALSE, message=FALSE, error=FALSE}
varImp(RF_model)
```
The five most important variables for prediction appear to be num_window, roll_belt, pitch_forearm, magnet_dumbbell_y, and magnet_dumbbell_z.

#### 4. Validate Random Forest model

```{r warning=FALSE, message=FALSE, error=FALSE}
testing_predict <- predict(RF_model,myTesting)
confusionMatrix(testing_predict,myTesting$classe)
```
When applying the trained Random Forest model to the validation set, the accuracy is 99.88%. No Class A false negatives were predicted, and only 6 false positives were.

#### 5. Predict 20 test cases

```{r warning=FALSE, message=FALSE, error=FALSE}
Test_Answers = predict(RF_model, newdata=testing_clean)
print(Test_Answers)
```

#### 6. References used

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.
